Storm surge modelling in the Strait of Georgia

### 2014-07-29 RoughWork/manuscript.tex
Added references for high tides.

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through comparisons of tidal elevation harmonics, tidal current ellipses, and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed, finding that the surge entering the domain from the Pacific Ocean contributes most significantly. In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas. Surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca. 
%needs a little more work.
\end{abstract}

@@

%some general intro paragraph here. Need something about SoG size, geography, islands, sills, rivers, tides....
The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait.

%my attempt at physical oceanopgraphy in SoG- Susan feel free to edit.
Several features related to the physical oceanopgraphy of this region render it a challenge to model numerically. The outflow from the Fraser River in the late spring is the primary source fresh water within the Strait of Georgia and leads to large vertical density gradients in the summer months. Additionally, strong tidal mixing through the San Juan and Gulf Islands produces horizontal density gradients that separate the saline waters of the Strait of Juan de Fuca and the fresher water of the Strait of Georgia. It is important to accurately represent the vertical mixing in this region as it sets the rate of export of fresh water through estuarine circulation. The strong and fast tidal currents through the narrow channels of the north, such as Discovery Passage and Johnstone Strait, pose a particular challenge for numerical models. Yet, several modelling efforts for this region exist, ranging in choices of grid structure, mixing parametrizations, and domain size and extent.

%overview oF SoG/Salish Sea models
Early work by \citet{crean1988mathematical,crean1988numerical} and \citet{stronach1993update} resulted in a series of models covering the Strait of Juan de Fuca, the Strait of Georgia and Puget Sound using a finite difference discretization of the equations of motion. A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinc version of their model examined residual circulation and fresh water flushing rates \citep{stronach1993update, marinone1996model}. This early work pointed to an important connection between the model's bottom friction parameterization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, especially in the area of Haro Strait, was necessary in order to adequately reproduce the observed tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction. \citet{foreman2004m} used data assimilation techinques to study M$_2$ tidal dissipation without local changes to the bottom frction in a two-dimensional finite element model.

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia noting a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. \citet{yang2010multi} developed a model of southern Strait of Georgia and Puget Sound using volume techniques on an unstructured grid. \citet{sutherland2011model} employed the Regional Ocean Modeling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia. The passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  These studies have led to an improvement in understanding circulation properties in the Salish Sea and have highlighted important features that should be considered in future modelling work.

@@

Communities in the Strait of Georgia are partiuclarly susceptible to storm surges in the winter months when there is an increased propensity for storms and a higher mean sea level caused by southeasterly winds and geostrophic adjustment \citep{danard2003storm}. In the summer months, the wind direction changes to northwesterly and mean sea level is lower. In British Columbia waters, the largest tidal range is typically observed in the winter months during the perigean spring tide or ``King" tide, when the sun and moon are aligned with the Earth and the moon is at its perigee. As such, storm surges are typically observed between Novemeber and Feburary. Several historical examples have been documented by \citet{forseth2006adaptation}. 
%work on this

@@

%overview of paper: model configuration, model evaluation, storm surge hindcasts.
This paper is organized as follows: an overview of the model configuration and domain including a description of the ocean model we have employed  is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parameterizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assesed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

%A section about how we configured the model for SoG 
We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration to develop an ocean model for the Strait of Georgia and Salish Sea. NEMO is a highly modularized tool used for studying ocean physics, ocean-ice interactions, and the biogeochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and biogeochemical coupling make it a useful tool for studying the complex physics and biogeochemical interactions within the Strait of Georgia. This work focuses on validating the physical set up of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include biogeochemical coupling and data assimilation.

The modelled domain extends from the Strait of Juan de Fuca to Puget Sound to Johnstone Strait as shown in Figure \ref{fig:domain}. Bathymetry from the Cascadia physiography dataset \citep{haugerud1999digital} was smoothed to limit the difference in depth across grid cells. For model stability, additional smoothing at the Strait of Juan de Fuca western boundary was imposed to achieve constant depth across the first ten grid cells. As depicted in Figure \ref{fig:domain}, the numerical grid is rotated $29^{\circ}$ counter-clockwise of North in order to maintain computational efficiency since currents within the Strait of Georgia are mainly aligned with this rotated axis.

@@

In addition to the equations of motion, a prognostic equation for the sea surface height is solved at each time step. The inclusion of the sea surface height equation requires a fairly restrictive time step due to the presence of high speed surface gravity waves. As such, the split-explicit time stepping algorithm is employed, where the free surface and barotropic equations are solved with a smaller time step than that used for the other variables. The model time step is and barotropic time step are 10 s and 2 s respectively. Small vertical grid spacing and large vertical velocities in the Boundary Pass and Haro Strait region impose a fairly restricitive step through the vertical CFL condition.

@@

At the ocean surface, meteorological conditions and fresh water input from rivers are included. The ocean surface is forced with momentum and heat fluxes from a 33-km global atmospheric reforcasting model suitable for use in ocean modelling \citep{smith2013new}. Forecasts from the period of 2002-2011 are available. Additionally, the inverse barometric effect of the atmospheric pressure is included, an important consideration in storm surge modelling.

River input provides a significant volume of freshwater to the Salish Sea and can influence stratification, circulation and primary productivity. However, most rivers in the domain are not gauged so parameterizations were required to represent river flow. \citet{morrison2011rivers} provides a method for estimating freshwater runoff in the Salish Sea region based on precipitation. Monthly runoff volumes for each watershed for each year from 1970 to 2012 were acquired from \citet{morrison2011rivers}, as well as monthly averages.

@@

The monthly flow from each river was input as a point source in the three grid points closest to the surface at the model point closest to the mouth of each river. Incoming water was assumed to be fresh and at surface temperature. A total of 150 rivers were parameterised by this method and their locations are indicated by the blue dots in Figure \ref{fig:domain}.

Initial conditions for temperature and salinity were taken from a CTD cast in the middle Strait of Georgia taken in Sept 2002 \citep{pawlowiczetal2007}.  Conditions were initially uniform horizontally and velocity was initialized at zero. The model was spun up for a 15.5 months from the initial conditions above, starting Sep 16, 2002, using atmospheric forcing from 2002-2003, climatological temperature and salinity and sea surface height at the boundaries, with tides and climatological river output.  All storm surge runs were started three days prior to the event of interest with zero initial velocities and sea surface height and a stratification profile from model spin up. The modelled sea surface height adjusted to forcing in less than one day.

@@

\subsection{Tidal evaluation}
The model was initially evaluated qualitatively by comparing patterns of tidal amplitude and phase to results from \citep{foreman1995tidal}. For example, the amphidromic dome around Victoria was produced in the M$_2$ results, as well as the monotonic increase in $K_1$ amplitude moving northwards along the Strait of Georgia. Initially, the Johnstone Strait boundary was closed, but modelled M$_2$ amplitudes were too small compared to measured amplitudes... TBC

@@

\section{Storm Surge Hindcasts}\label{sec:storm}
%outline: 
%1. factors contributing to storm surge with focus on Feb 2006 event
%2. Strong wind event but no surge (maybe Dec 2006 or Nov 2009)
%3. High resolution wind (hopefully Dec 2012)
A series of storm surge hindcasts is presented next. Storm surge simulations were started from rest at least three days prior to the event of interest to allow for adjustment of model currents. Stratification conditions from model spin-up at a time close to the event (within 5 days) were chosen to initialize the temperature and salinity fields. Boundary conditions include atmospheric forcing and fresh water input from rivers at the surface, sea surface height anomaly from observations at the open boundary, and tidal forcing with eight tidal constituents. Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height predictions due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can be up to 40 cm (check) during times of high tidal range, and in particular, during the storm surge season Nov-Feb. As such, storm surge model predictions were corrected by accounting for this error. Additionally, since the sea surface height is calculated about a zero reverence level, the modelled water level is determined by adding a long term average of mean sea level at the specified location to the modelled sea surface height and removing a background anomaly caused by topographic pressure gradients since the atmospheric reforecasting model uses a terrain following coordinate system. (Nancy: I think we need a better way of adjusting come publication time).

Model results have been compared with observations by calculating both the total modelled water level and the modelled residual. The modelled residual is calculated as the difference in model output between a simulation with all forcing conditions and a simulation forced only with tides and rivers. The observed water level is taken from observations by Fisheries and Oceans Canada (citation) and the observed residual is the observations minus tidal predictions. Tidal predictions are calculated using t\_tide \citep{pawlowicz2002classical}, a MATLAB tool that calculates tidal harmonics from a given time series, in this cae, from the year prior surge event on interest. Additionally, several simulations with varying forcing conditions such as removing atmospsheric forcing or removing the sea surface height anomaly at the open boundaries have been completed.

Several significiant surge or wind events have been considered in detail, as well as an array of hindcasts to asses model ability in a statistical measure. These cases have been used to asses the model's skill in reproducing water level height and surge amplitude and timing during storm events. In addition, the contribution to the surge due to several forcing conditions, such as atmospheric forcing and open boundary forcing, has been assessed. We have also examined the spatial variabilty of the surge as it propagates through the domain and have considered a case with high-resolution atmospheric forcing.

%First, a study of a large surge event on Feb 4, 2006 is presented. This event caused siginificant damage and flooding in Boundary Bay, a community south of Richmond, (citation) and has been chosen as a case study to demonstrate the model's skill in reproducing a storm surge and to quantitatively assess forcing conditions that are important in storm surge hindcasting. Second, an extreme wind event on Dec 15, 2006 is examined. This event produced winds over (x) (citation) but did not result in significant flooding because the storm did not occur at high tide, however, it did result in siginificant property damage due to high winds and waves. This event has been used to understand the spatial and temporal variability of the surge propagation. Next, a large wind event on Nov 18, 2009 that did not result in a siginificant surge or flooding is presented. This case was choosen to identify the model's capability in handling large wind events that do not result in high water levels. Next, a case with high resolution atmospheric forcing for a recent surge on Dec 17, 2012 is discussed. Lastly, these cases presented in detail will be combined with $x$ additional runs for statistical evaluation of model performance.

@@

On Feb 4, 2006, a strong storm system with winds over 75 km/hr coinciding with an unsually high tide led to siginificant flooding in coastal comminuties of the Strait of Georgia \citep{romanowski2010storm}. The maximum water level observed at Point Atkinson on this day was 5.49 m, which is 0.8 m higher than the predicted tides, highlighting the significance of this event as it resulted in a large anomaly and also reached water levels close to the highest level ever recorded (5.61 m). A set of simulations from Feb 1 to 8, 2006 under varying forcing conditions has been completed in order to determine the siginicance of each factor contributing to the surge.

The model's ability to reproduce this event is demonstrated in Figure \ref{fig:feb2006} which compares observations and corrected model output at Point Atkinson, Victoria, Patricia Bay, and Campbell River. The timing of maximum water level matches well between the observations and the model at all four locations, the most siginificant contribution to high water level being from the tides. There is also a good match in amplitude of the water level maximum at Point Atkinson and Campbell River, with a relative error less than 4\%. There is only a 7.2 cm difference between observed and modelled water level elevation at Point Atkinson and a 20.9 cm difference at Campbell River. Tthe error at Victoria and Patricia Bay is slightly more siginificant, reaching 9.6\% and 6.3\% respectively (close to 30 cm in each case), partly due to error in tidal amplitudes as discussed above. The precise location and magnitude of the M$_2$ amphidrome around Victoria is very sensitive to model parameters such as bottom friction.

@@

The contribution to the surge due to atmospheric forcing appears to be geographically dependent, its removal leading to a drop of about 5 cm in the maximum surge at Point Atkinson and 10 cm at Campbell River. When atmospheric forces are neglected the timing of the maximum surge is earlier at Point Atkinson but later at Campbell River. Campbell River exhibits a strong response to atmopsheric conditions during midday on Feb 4, as demonstrated in a two sharp spikes in both the blue and red curves arounf 12:00 hr The green curve, which does not include any atmopsheric focring, does not exhibit this behaviour. The result is a peak surge earlier than in day than the no atmospheric forcing case. This response is also noted in the observations at Campbell River in Figure \ref{fig:feb2006} where the maxmium surge occurs earlier than at the other locations. At Victoria and Patricia Bay, the surge amplitude is mainly unaffected by the atmospheric forcing. 
%maybe look at winds/pressure to see why Vic/Pat Bay aren't affected?. Plan: map of wind vectors from CGRF (not for paper, just for me in aiding with explanation)

These comparisons point to the importance of including, as a forcing condition at the open bounday, the surge entering the domain from the Pacific Ocean. Storms travelling over the Pacific Ocean torwads the west coast of the northern United States and Canada can induce elevated sea levels along the coastline which then enter the Salish Sea sytem through the Strait of Juan de Fuca. In a model domain that does not inlcude the Pacific Ocean, this effect must be inlcuded as a boundary condition. Previous simulations by \citet{murty1995storm} suggest that the inverese barometer effect and the surge entering the system through the Strait of Juan de Fuca contribute most significantly to storm surges in this domain.  A detailed assessment of the effect of the local atmospheric condtions and winds is included later where simulations have been performed with high resolution atmospheric forcing data.

@@

%Finally, an assessment of the importance of including tide-surge interactions is presented next. A simulation of the surge propagation without including any tidal forcing is compared with the modelled residual in Figure \ref{fig:tidesurge}. Recall, the modelled residual is defined as the difference between the sea surface height in a simulation with all forcing and a simulaltion with tides amd rivers only. A comparison between the modelled residual and a surge only run can determine the significance of the tide-surge interaction. This figure shows there is almost no difference between the simulations at all locations, rendering the tide-sure interaction negligible. In operational settings, running a surge only simulation and adding the predicted tides a posteriori could be more computationally efficient with only a small decrease in the accuracy of the prediction. 
% I'm not sure if removing a background state is really appropriate here. There could be some differences in the background states because of the tides(ie where fresh water moves to, etc). I also think it is more appropriate to use the atmospheric pressure to remove the background state...

>>>>

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hind casts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through comparisons of tidal elevation harmonics, tidal current ellipses, and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed, finding that the surge entering the domain from the Pacific Ocean contributes most significantly. In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas. Surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca. 
%needs a little more work.
\end{abstract}

@@

%some general intro paragraph here. Need something about SoG size, geography, islands, sills, rivers, tides....
The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait. It is on average 222 km in length and 28 km wide and its maximum depth is 420 m \citep{thomson1981oceanography}. At its southern end, the Strait of Georgia is connected to the Strait of Juan de Fuca through several passages, the largest being Haro Strait. The San Juan and Gulf Islands are scattered throughout these passages where complex bathymetry and steep topographic slopes induce vigorous tidal mixing. Many rivers feed fresh water into the Salish Sea system, the most significant source being from the Fraser River. Mixed diurnal and semi-diurnal tides are found throughout most of the Salish Sea but a semi-diurnal amphidrome is present nearby Victoria, where the tides are mainly diurnal. This system is home to several large coastal communities, such as Vancouver and Victoria, serves as a waterway for commercial traffic and recreational use, and supports a large and active fishing industry.

%my attempt at physical oceanography in SoG- Susan feel free to edit.
Several features related to the physical oceanography of this region render it a challenge to model numerically. The outflow from the Fraser River in the late spring is the primary source fresh water within the Strait of Georgia and leads to large vertical density gradients in the summer months. Additionally, strong tidal mixing through the San Juan and Gulf Islands produces horizontal density gradients that separate the saline waters of the Strait of Juan de Fuca and the fresher water of the Strait of Georgia. It is important to accurately represent the vertical mixing in this region as it sets the rate of export of fresh water through estuarine circulation. The strong and fast tidal currents through the narrow channels of the north, such as Discovery Passage and Johnstone Strait, pose a particular challenge for numerical models. Yet, several modelling efforts for this region exist, ranging in choices of grid structure, mixing parametrizations, and domain size and extent.

%overview oF SoG/Salish Sea models
Early work by \citet{crean1988mathematical,crean1988numerical} and \citet{stronach1993update} resulted in a series of models covering the Strait of Juan de Fuca, the Strait of Georgia and Puget Sound using a finite difference discretization of the equations of motion. A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinic version of their model examined residual circulation and fresh water flushing rates \citep{stronach1993update, marinone1996model}. This early work pointed to an important connection between the model's bottom friction parametrization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, especially in the area of Haro Strait, was necessary in order to adequately reproduce the observed tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction. \citet{foreman2004m} used data assimilation techniques to study M$_2$ tidal dissipation without local changes to the bottom friction in a two-dimensional finite element model.

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia noting a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. \citet{yang2010multi} developed a model of southern Strait of Georgia and Puget Sound using volume techniques on an unstructured grid. \citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia. The passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  These studies have led to an improvement in understanding circulation properties in the Salish Sea and have highlighted important features that should be considered in future modelling work.

@@

Communities in the Strait of Georgia are particularly susceptible to storm surges in the winter months when there is an increased propensity for storms and a higher mean sea level caused by southeasterly winds and geostrophic adjustment \citep{danard2003storm}. In the summer months, the wind direction changes to northwesterly and mean sea level is lower. In British Columbia waters, the largest tidal ranges occur in June and December at the summer and winter solstices \citep{thomson1981oceanography}. The highest water levels are typically observed in the winter around the perigean spring tide or ``King" tide, when the sun and moon are aligned with the Earth and the moon is at its perigee. As such, storm surges are typically observed between November and February. Several historical examples have been documented by \citet{forseth2006adaptation}. 
%work on this
%something about fetch

@@

%overview of paper: model configuration, model evaluation, storm surge hind casts.
This paper is organized as follows: an overview of the model configuration and domain including a description of the ocean model we have employed  is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hind casts is also assessed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

%A section about how we configured the model for SoG 
We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration to develop an ocean model for the Strait of Georgia and Salish Sea. NEMO is a highly modularized tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on validating the physical set up of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation.

The modelled domain extends from the Strait of Juan de Fuca to Puget Sound to Johnstone Strait as shown in Figure \ref{fig:domain}. Bathymetry from the Cascadia physiography data set \citep{haugerud1999digital} was smoothed to limit the difference in depth across grid cells. For model stability, additional smoothing at the Strait of Juan de Fuca western boundary was imposed to achieve constant depth across the first ten grid cells. As depicted in Figure \ref{fig:domain}, the numerical grid is rotated $29^{\circ}$ counter-clockwise of North in order to maintain computational efficiency since currents within the Strait of Georgia are mainly aligned with this rotated axis.

@@

In addition to the equations of motion, a prognostic equation for the sea surface height is solved at each time step. The inclusion of the sea surface height equation requires a fairly restrictive time step due to the presence of high speed surface gravity waves. As such, the split-explicit time stepping algorithm is employed, where the free surface and barotropic equations are solved with a smaller time step than that used for the other variables. The model time step is and barotropic time step are 10 s and 2 s respectively. Small vertical grid spacing and large vertical velocities in the Boundary Pass and Haro Strait region impose a fairly restrictive step through the vertical CFL condition.

@@

At the ocean surface, meteorological conditions and fresh water input from rivers are included. The ocean surface is forced with momentum and heat fluxes from a 33-km global atmospheric reforecasting model suitable for use in ocean modelling \citep{smith2013new}. Forecasts from the period of 2002-2011 are available. Additionally, the inverse barometric effect of the atmospheric pressure is included, an important consideration in storm surge modelling.

River input provides a significant volume of freshwater to the Salish Sea and can influence stratification, circulation and primary productivity. However, most rivers in the domain are not gauged so parametrizations were required to represent river flow. \citet{morrison2011rivers} provides a method for estimating freshwater runoff in the Salish Sea region based on precipitation. Monthly runoff volumes for each watershed for each year from 1970 to 2012 were acquired from \citet{morrison2011rivers}, as well as monthly averages.

@@

The monthly flow from each river was input as a point source in the three grid points closest to the surface at the model point closest to the mouth of each river. Incoming water was assumed to be fresh and at surface temperature. A total of 150 rivers were parametrized by this method and their locations are indicated by the blue dots in Figure \ref{fig:domain}.

Initial conditions for temperature and salinity were taken from a CTD cast in the middle Strait of Georgia taken in Sept 2002 \citep{pawlowiczetal2007}.  Conditions were initially uniform horizontally and velocity was initialized at zero. The model was spun up for a 15.5 months from the initial conditions above, starting Sept 16, 2002, using atmospheric forcing from 2002-2003, climatological temperature and salinity and sea surface height at the boundaries, with tides and climatological river output.  All storm surge runs were started three days prior to the event of interest with zero initial velocities and sea surface height and a stratification profile from model spin up. The modelled sea surface height adjusted to forcing in less than one day.

@@

\subsection{Tidal evaluation}
The model was initially evaluated qualitatively by comparing patterns of tidal amplitude and phase to results from \citep{foreman1995tidal}. For example, the amphidrome around Victoria was produced in the M$_2$ results, as well as the monotonic increase in $K_1$ amplitude moving northwards along the Strait of Georgia. Initially, the Johnstone Strait boundary was closed, but modelled M$_2$ amplitudes were too small compared to measured amplitudes... TBC

@@

\section{Storm Surge Hind casts}\label{sec:storm}
%outline: 
%1. factors contributing to storm surge with focus on Feb 2006 event
%2. Strong wind event but no surge (maybe Dec 2006 or Nov 2009)
%3. High resolution wind (hopefully Dec 2012)
A series of storm surge hind casts is presented next. Storm surge simulations were started from rest at least three days prior to the event of interest to allow for adjustment of model currents. Stratification conditions from model spin-up at a time close to the event (within 5 days) were chosen to initialize the temperature and salinity fields. Boundary conditions include atmospheric forcing and fresh water input from rivers at the surface, sea surface height anomaly from observations at the open boundary, and tidal forcing with eight tidal constituents. Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height predictions due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can be up to 40 cm (check) during times of high tidal range, and in particular, during the storm surge season Nov-Feb. As such, storm surge model predictions were corrected by accounting for this error. Additionally, since the sea surface height is calculated about a zero reverence level, the modelled water level is determined by adding a long term average of mean sea level at the specified location to the modelled sea surface height and removing a background anomaly caused by topographic pressure gradients since the atmospheric reforecasting model uses a terrain following coordinate system. (Nancy: I think we need a better way of adjusting come publication time).

Model results have been compared with observations by calculating both the total modelled water level and the modelled residual. The modelled residual is calculated as the difference in model output between a simulation with all forcing conditions and a simulation forced only with tides and rivers. The observed water level is taken from observations by Fisheries and Oceans Canada (citation) and the observed residual is the observations minus tidal predictions. Tidal predictions are calculated using t\_tide \citep{pawlowicz2002classical}, a MATLAB tool that calculates tidal harmonics from a given time series, in this cae, from the year prior surge event on interest. Additionally, several simulations with varying forcing conditions such as removing atmospheric forcing or removing the sea surface height anomaly at the open boundaries have been completed.

Several significant surge or wind events have been considered in detail, as well as an array of hind casts to asses model ability in a statistical measure. These cases have been used to asses the model's skill in reproducing water level height and surge amplitude and timing during storm events. In addition, the contribution to the surge due to several forcing conditions, such as atmospheric forcing and open boundary forcing, has been assessed. We have also examined the spatial variability of the surge as it propagates through the domain and have considered a case with high-resolution atmospheric forcing.

%First, a study of a large surge event on Feb 4, 2006 is presented. This event caused significant damage and flooding in Boundary Bay, a community south of Richmond, (citation) and has been chosen as a case study to demonstrate the model's skill in reproducing a storm surge and to quantitatively assess forcing conditions that are important in storm surge hind casting. Second, an extreme wind event on Dec 15, 2006 is examined. This event produced winds over (x) (citation) but did not result in significant flooding because the storm did not occur at high tide, however, it did result in significant property damage due to high winds and waves. This event has been used to understand the spatial and temporal variability of the surge propagation. Next, a large wind event on Nov 18, 2009 that did not result in a significant surge or flooding is presented. This case was chosen to identify the model's capability in handling large wind events that do not result in high water levels. Next, a case with high resolution atmospheric forcing for a recent surge on Dec 17, 2012 is discussed. Lastly, these cases presented in detail will be combined with $x$ additional runs for statistical evaluation of model performance.

@@

On Feb 4, 2006, a strong storm system with winds over 75 km/hr coinciding with an unusually high tide led to significant flooding in coastal communities of the Strait of Georgia \citep{romanowski2010storm}. The maximum water level observed at Point Atkinson on this day was 5.49 m, which is 0.8 m higher than the predicted tides, highlighting the significance of this event as it resulted in a large anomaly and also reached water levels close to the highest level ever recorded (5.61 m). A set of simulations from Feb 1 to 8, 2006 under varying forcing conditions has been completed in order to determine the significance of each factor contributing to the surge.

The model's ability to reproduce this event is demonstrated in Figure \ref{fig:feb2006} which compares observations and corrected model output at Point Atkinson, Victoria, Patricia Bay, and Campbell River. The timing of maximum water level matches well between the observations and the model at all four locations, the most significant contribution to high water level being from the tides. There is also a good match in amplitude of the water level maximum at Point Atkinson and Campbell River, with a relative error less than 4\%. There is only a 7.2 cm difference between observed and modelled water level elevation at Point Atkinson and a 20.9 cm difference at Campbell River. The error at Victoria and Patricia Bay is slightly more significant, reaching 9.6\% and 6.3\% respectively (close to 30 cm in each case), partly due to error in tidal amplitudes as discussed above. The precise location and magnitude of the M$_2$ amphidrome around Victoria is very sensitive to model parameters such as bottom friction.

@@

The contribution to the surge due to atmospheric forcing appears to be geographically dependent, its removal leading to a drop of about 5 cm in the maximum surge at Point Atkinson and 10 cm at Campbell River. When atmospheric forces are neglected the timing of the maximum surge is earlier at Point Atkinson but later at Campbell River. Campbell River exhibits a strong response to atmospheric conditions during midday on Feb 4, as demonstrated in a two sharp spikes in both the blue and red curves around 12:00 hr The green curve, which does not include any atmospheric coring, does not exhibit this behaviour. The result is a peak surge earlier than in day than the no atmospheric forcing case. This response is also noted in the observations at Campbell River in Figure \ref{fig:feb2006} where the maximum surge occurs earlier than at the other locations. At Victoria and Patricia Bay, the surge amplitude is mainly unaffected by the atmospheric forcing. 
%maybe look at winds/pressure to see why Vic/Pat Bay aren't affected?. Plan: map of wind vectors from CGRF (not for paper, just for me in aiding with explanation)

These comparisons point to the importance of including, as a forcing condition at the open boundary, the surge entering the domain from the Pacific Ocean. Storms travelling over the Pacific Ocean towards the west coast of the northern United States and Canada can induce elevated sea levels along the coastline which then enter the Salish Sea system through the Strait of Juan de Fuca. In a model domain that does not include the Pacific Ocean, this effect must be included as a boundary condition. Previous simulations by \citet{murty1995storm} suggest that the inverse barometer effect and the surge entering the system through the Strait of Juan de Fuca contribute most significantly to storm surges in this domain.  A detailed assessment of the effect of the local atmospheric conditions and winds is included later where simulations have been performed with high resolution atmospheric forcing data.

@@

%Finally, an assessment of the importance of including tide-surge interactions is presented next. A simulation of the surge propagation without including any tidal forcing is compared with the modelled residual in Figure \ref{fig:tidesurge}. Recall, the modelled residual is defined as the difference between the sea surface height in a simulation with all forcing and a simulation with tides and rivers only. A comparison between the modelled residual and a surge only run can determine the significance of the tide-surge interaction. This figure shows there is almost no difference between the simulations at all locations, rendering the tide-sure interaction negligible. In operational settings, running a surge only simulation and adding the predicted tides a posteriori could be more computationally efficient with only a small decrease in the accuracy of the prediction. 
% I'm not sure if removing a background state is really appropriate here. There could be some differences in the background states because of the tides(ie where fresh water moves to, etc). I also think it is more appropriate to use the atmospheric pressure to remove the background state...

Finally, an assessment of the importance of including the tide-surge interaction is presented next. A simulation of the surge propagation without including any tidal forcing is compared with the modelled residual in Figure \ref{fig:tidesurge}.  Recall, the modelled residual is defined as the difference between the sea surface height in a simulation with all forcing and a simulation with tides and rivers only. A comparison between the modelled residual and a surge-only run can determine the significance of the tide-surge interaction. The most significant differences occur at Point Atkinson where the maximum modelled residual is 7 cm lower than the maximum of the surge-only case.  In contrast, the tide-surge interaction at Victoria is hardly noticeable. Indeed, a spatial map of the modelled residual and surge-only difference (not shown here) indicates that the tide-surge interaction is more pronounced after the San Juan and Gulf Islands. The passages between these islands are known for vigorous tidal mixing, which may have an effect on the surge propagation. However, the effect is a relatively small underestimate of the surge elevation and no changes to the timing of the surge. So, in operational settings, performing a surge-only simulation and adding the predicted tides a posteriori could be more computationally efficient with only a small decrease in the accuracy of the prediction.
